### 2023-03-15 paper.tex
Updates from Overleaf

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_AB.png}\\
    \end{center}
    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals.}\\
    \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
    Two cells cover differing parts of the interval with one read each.
    If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells.
    \textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained.
    To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
    \label{fig:smoothres}
\end{figure*}

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=0.85\columnwidth]{figures/Fig_sliding.png}
    \end{center}
    \caption{\small \textbf{Finding variably methylated regions.}\\
    The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
    Then, the variance of these values is calculated (each point represents one of the overlapping windows), a threshold (dashed line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
    Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions'' (VMRs).}
    \label{fig:vmr}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\columnwidth]{figures/Fig_DMRdetection.png}
    \caption{\small \textbf{Identification of differentially methylated regions.}\\
    The chromosome is divided into windows overlapping by a small and fixed size.
    For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
    Then the t-statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
    Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t-statistic above or below the threshold.
    Windows exceeding either threshold are merged if they overlap, yielding the differentially methylated regions (DMRs).}
    \label{fig:dmrscan}
\end{figure}

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=.95\columnwidth]{figures/Fig_correlation.png}
    \end{center}
    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.\\
    \textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
    \textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
    Promoters are defined as intervals \textpm2~kb around the TSS.
    \textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
    Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
    Data from \citet{kremer_scnmt}.}
    \label{fig:correlation}
\end{figure}

\subsection{Our proposed methods improve identification of cell types}

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=0.8\textwidth]{figures/Fig_benchmark.png}
    \end{center}
    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.\\
    \textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
    \textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (B) or conventional methods based on genome tiling (C).
    \textbf{(D, E)} Mean neighbor score \textbf{(D)} obtained after analysing single-cell methylomes with different combinations of methods.
    Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA) or using conventional methods (PCA on lightly imputed methylation percentages, based on \citep{luo2017single}).
    The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells \textbf{(A)}.
    }
    \label{fig:score}
\end{figure*}

@@

To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed in a neighborhood consisting of cells with the same cell type in 15-dimensional PC space ("neighbor score", for details see Methods).
Since scBS protocols are costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them with different method combinations, and calculated the mean neighbor score for each (Figure \ref{fig:score}D,E).
The results demonstrate that the forebrain cell clusters are more cleanly separated when using our proposed methods.
This performance gain was observed also in smaller data sets, although cell cluster separation generally becomes more difficult in those.
Both the use of VMRs instead of genomic tiles, as well as the use of the shrunken mean of the residuals over methylation averages improved cell cluster separation.

@@

\phantomsection
\subsection{Detecting DMRs between oligodendrocytes and neural stem cells}

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=.7\textwidth]{figures/Fig_DMRs.png}
    \end{center}
    \caption{\small \textbf{Detection of differentially methylated regions between oligodendrocytes and neural stem cells.}\\
    \textbf{(A)} DMRs detected between 58 oligodendrocytes (ODs) and 130 neural stem cells (NSCs) from \citep{kremer_scnmt} (top) and DMRs detected in the same data with randomly permuted cell labels (bottom, used to estimate FDR and determine adjusted p-values).
    \textbf{(B)} Enrichment of GO terms associated with DMRs lowly methylated in ODs (top) or NSCs (bottom).
    Depicted are the top 5 GO terms of the "Biological Process" (GO:BP) and "Cellular Component" (GO:CC) GO category, and their binomial test p-value and enrichment, as reported by GREAT \citep{mclean2010great}.
    \textbf{(C)} Mean methylation of NSCs and ODs at an exemplary DMR.
    Each point corresponds to a cell, black lines denote the median.
    \textbf{(D)} Detailed view of the DMR (yellow dashed rectangle) from \textbf{C} in pseudo-bulk samples consisting of NSCs or ODs.
    Vertical bars represent CpG sites.
    }
    \label{fig:dmr}
\end{figure*}

Next, we used the single-cell multi-omics data set \citep{kremer_scnmt} to evaluate our DMR detection approach.
We first selected two cell populations from the healthy murine ventricular-subventricular zone: neural stem cells (NSCs, 130 cells) and oligodendrocytes (58 cells).
Then we used the method depicted in Fig.~\ref{fig:dmrscan} to identify DMRs between these two cell types (Fig.~\ref{fig:dmr}).
To estimate the false discovery rate and obtain adjusted p-values, we repeated the DMR detection approach on a permutation of the same data set, generated by shuffling cell labels.
DMRs detected in this manner showed more modest methylation differences, and as a result many DMRs detected in the original data were assigned low adjusted p-values (Fig.~\ref{fig:dmr}A).
Gene ontology (GO) enrichment with GREAT \citep{mclean2010great} revealed that DMRs lowly methylated in oligodendrocytes are located near genes involved in myelination, the main function of oligodendrocytes.
Similarly, DMRs specifically demethylated in NSCs occur near genes involved in stem cell population maintenance.
This demonstrates that our DMR detection approach is able to identify biologically meaningful DMRs, even in scBS data sets of modest cell number.

Fig.~\ref{fig:dmr}C-D depicts an exemplary DMR, located at the gene body of myelin-basic protein (Mbp), the major component of myelin that is essential for myelination of neuronal axons \citep{mbp}.
Our results suggest that oligodendrocyte-specific gene expression of Mbp is supported by low methylation at the detected DMR.

@@

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
    \end{center}
    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}\\
        \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimised for efficient access to genomic intervals.
        \texttt{scbs} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
        These cells can be discarded with \texttt{scbs filter}.
        To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
        The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{scbs scan}.
        The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
        Identified groups of cells (e.~g. clusters) can then be scanned for DMRs with \texttt{scbs diff}.
    }
    \label{fig:workflow}
\end{figure*}

We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{fig:workflow}).
The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark}, methylpy \citep{methylpy} or biscuit \citep{biscuit}.
For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
\texttt{scbs prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or custom user-defined formats.

@@

After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{scbs scan} command.
This produces a BED file that lists the genome coordinates of VMRs.
To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells.
The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.

Finally, after annotation and exploration of the data set, the user may specify two groups of cells for DMR detection with \texttt{scbs diff}.
This command produces a BED file listing the genome coordinates, adjusted p-values, as well as several other metrics associated with DMRs.
These DMRs may then be associated with nearby genes, or used for GO enrichment with tools such as GREAT \citep{mclean2010great} to enable functional interpretation.

@@

Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably methylated regions (VMRs).
By re-analyzing two published data sets, we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for challenging data sets comprising cellular sub-states and lineage transitions.

To furthermore aid interpretation of scBS data, we developed and implemented the first single-cell DMR detection algorithm.
FDR estimation via permutation allows us to report statistical significance of each DMR.
% maybe here discuss pseudoreplication issues?
Applying our approach to NSCs and oligodendrocytes demonstrated that the obtained DMRs locate near meaningful loci associated with cell-type specific functions.
Thus, our method aids biological interpretability by pinpointing the regions that differ between groups of cells and hence have a putative regulatory role.

@@

In addition, by pinpointing exact regions of interest, the VMR and DMR functionalities provide useful input for genomics-style analyses.
Depending on the research question at hand, individual DMRs can be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology enrichment or motif enrichment.

@@

\begin{figure}[h]
    \begin{center}
        \includegraphics[width=0.9\columnwidth]{figures/SFig_benchmark.png}
    \end{center}
    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
    Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
    }
    \label{fig:score_luo}
\end{figure}

\begin{figure}[h]
    \begin{center}
        \includegraphics[width=0.9\columnwidth]{figures/SFig_benchmark.png}
    \end{center}
    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.\\
    Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
    }
    \label{fig:score_luo}
\end{figure}

@@

S.\ A.\ and L.\ P.\ M.\ K.\ acknowledge funding by the Klaus Tschira Foundation.
L.\ P.\ M.\ K., L.\ K., S.\ C.\ and A.\ M.-V.\ acknowledge funding from the European Commission via ERC grant 771376.
We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.

\phantomsection
\section{Methods (note to Simon: didnt go through this yet)}  % todo rm

@@

Whenever a region is not covered by any read in a cell, the corresponding data entry in the input data matrix for PCA will be missing.
The standard approach to calculate PCA, commonly done using the IRLBA algorithm \citep{Baglama2005}, is not suited to deal with missing data.
We circumvent this issue by simply using the PCA itself to impute the missing value.

@@

Our iterative imputation strategy is now simply the following: we first replace all missing values in the row-centered input matrix $A$ with zeroes and perform the (truncated) PCA.
Then, we use the PCA predictions for the missing values, i.e., the $a_{ij}=\sum_r x_{ir} r_{jr}$, as refined stand-ins for the missing values in $A$ and run PCA once more.
This can now be iterated until convergence.

\subsection{Re-analysis of published scBS datasets}
To re-analyze scBS data sets by \citet{kremer_scnmt} and \citet{luo2017single}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 2\%$).
For all regions (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of the residuals were obtained with \texttt{scbs matrix}.

Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.

>>>>

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_AB.png}\\
    \end{center}
    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals:}
    \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified.
    Two cells cover differing parts of the interval with one read each.
    If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells.
    \textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained.
    To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
    \label{fig:smoothres}
\end{figure*}

@@

\begin{figure}
    \includegraphics[width=\columnwidth]{figures/Fig_sliding.png}
    \caption{\small \textbf{Finding variably methylated regions (VMRs):} The chromosomes are divided up into overlapping windows (first five shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.~\ref{fig:smoothres}B.
    Then, the variance of these values is calculated (each point represents one of the overlapping windows), a threshold (dashed line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold.
    Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions''.}
    \label{fig:vmr}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\columnwidth]{figures/Fig_DMRdetection.png}
    \caption{\small \textbf{Identification of differentially methylated regions (DMRs):}
    The chromosome is divided into windows overlapping by a small and fixed size.
    For each window, the cell’s methylation values are calculated as described and as depicted in Fig.~ \ref{fig:smoothres}B.
    Then the t-statistic of these values is obtained as a measure of methylation difference (each point represents one of the windows).
    Upper and lower thresholds (dashed lines) are chosen such that a defined quantile of windows have a t-statistic above or below the threshold.
    Windows exceeding either threshold are merged if they overlap, yielding the ``differentially methylated regions``.}
    \label{fig:dmrscan}
\end{figure}

@@

\begin{figure}
    \begin{center}
    \includegraphics[width=.95\columnwidth]{figures/Fig_correlation.png}
    \end{center}
    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
    \textbf{(A)} Distribution of Pearson correlations between gene expression and promoter methylation (black) and gene expression and methylation of the nearest VMR (red).
    \textbf{(B)} Correlation of each gene's RNA expression values to methylation (shrunken mean of residuals) at promoters (y-axis), or to methylation of the nearest VMR (x-axis).
    Promoters are defined as intervals \textpm2~kb around the TSS.
    \textbf{(C)} Mean methylation near the gene Htra1 (ENSMUSG00000006205).
    Cells are assigned to 4 groups based on Htra1 expression (group 0: cells with no Htra1 expression, groups 1-3: cells that express Htra1, divided into three equally large groups with group 3 having the highest expression).
    Data from \citet{kremer_scnmt}.}
    \label{fig:correlation}
\end{figure}

\subsection{Effects of our methods on cell type identification}

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=0.8\textwidth]{figures/Fig_benchmark.png}
    \end{center}
    \caption{\small \textbf{Benchmark of our methods on single-cell multi-omic data of cells of the murine forebrain}.
    \textbf{(A)} Cell labels based on clustering of single-cell transcriptomes from \citet{kremer_scnmt}.
    \textbf{(B, C)} Exemplary UMAPs obtained when analyzing the data set with our proposed methods (B) or conventional methods based on genome tiling (C).
    \textbf{(D, E)} Mean neighbor score \textbf{(D)} obtained after analysing single-cell methylomes with different combinations of methods.
    Either 100~kb genomic tiles or VMRs were subjected to PCA and UMAP.
    Tiles or VMRs were pre-processed as proposed in this work (shrunken mean of residuals subjected to iterative PCA, labeled "pre-processing: ours") or using conventional methods (PCA on lightly imputed methylation percentages, based on \citet{luo2017single}).
    The full 1566-cell data set was sub-sampled to simulate smaller data sets (x axis).
    A higher neighbor score implies better separation of cell types inferred from single-cell transcriptomes of the same cells (A).
    }
    \label{fig:score}
\end{figure*}

@@

To quantify this performance gain in a more rigorous manner, we used a score that quantifies whether cells were placed in a neighborhood consisting of cells with the same cell type in 15-dimensional PC space ("neighbor score", for details see Methods).
Since scBS protocols are costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
To simulate smaller data sets, we sub-sampled the 1566-cell data set into smaller data sets, analyzed them with different method combinations, and calculated the mean neighbor score for each (Figure \ref{fig:score}D,E).
The results demonstrate that the forebrain cell clusters are more cleanly separated when using our proposed methods.
This performance gain was observed also in smaller data sets, although cell cluster separation generally becomes more difficult in those.
Both the use of VMRs instead of genomic tiles, as well as the use of the mean shrunken residuals over methylation averages improved cell cluster separation.

@@

\section{note to Simon: this is how far I got}   % todo rm

\phantomsection
\subsection{Detecting DMRs between oligodendrocytes and neuroblasts}

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=.8\textwidth]{figures/Fig_DMRs.png}
    \end{center}
    \caption{\small \textbf{identified neuroblast and oligodendrocyte DMRs have a putative regulatory role:} \textbf{(A)} Difference in methylation between oligodendrocytes and neuroblasts of DMRs plotted against the number of cells contained in a DMR.
Colored by p-value.
\textbf{(B)} Mean methylation of cells in the DMR covering MBP.
Line indicates median.
\textbf{(C)} Mean methylation of reads at individual sites along the DMR covering MBP.
\textbf{(D)} GREAT \citet{mclean2010great} was used for GO-term analysis.
The left side shows GO-terms for genes near the 3000 most significant neuroblast DMRs sorted by GO-enrichment p-value and t-statistic (green arrow) and the right side for oligodendrocyte DMRs (yellow arrow) respectively.
    }
    \label{fig:dmr}
\end{figure*}

Next, we used the single-cell multi-omics data set \citep{kremer_scnmt} to assess the ability of our DMR detection approach.
We first selected two cell populations from the murine ventricular-subventricular zone, neuroblasts and oligodendrocytes, and then used the method depicted in Fig.~\ref{fig:dmrscan} to identify regions with differing methylation.
Fig.~\ref{fig:dmr} gives an overview of all identified DMRs and the relation of the adjusted p-value to the number of cells included in a DMR, and the difference in methylation between the cell types.
Logically, if a DMR contains few cells, the methylation difference between cell types must be higher to obtain a low (and thus more significant) p-value.
A low p-value is maintained, although the methylation difference decreases if the number of cells in the DMR increases.

An oligodendrocyte DMR stretching across Mbp was chosen to assess the methylation patterns of an individual DMR (Fig.~\ref{fig:dmr}B-C).
Mbp is highly expressed in oligodendrocytes as it is essential for the myelination of neuronal axons.
The DMR is highly methylated in neuroblasts and lowly methylated in oligodendrocytes (Fig.~\ref{fig:dmr}B-C), including more substantial demethylation in oligodendrocytes at the position of Mbp (Fig.~\ref{fig:dmr}C).
Low methylation at this locus supports the expression of Mbp.

Nevertheless, this is not the only indicator of the biological relevance of discovered DMRs.
GO-term enrichment shows that neuroblast DMRs occur near genes involved in the regulation of neurogenesis, neuron differentiation, and the positive regulation of neural precursor cell proliferation, all of which are essential for neuroblasts (Fig.~\ref{fig:dmr}D).
On the other side, oligodendrocyte-defining DMRs are located near genes with implications for myelin sheath, myelination, and axon ensheathment and thus are linked to oligodendrocytes.
Therefore, DMRs occur in the proximity of genes associated with oligodendrocyte or neuroblast function, respectively and are implied to regulate gene expression and contribute to fundamental cell type-specific function, which is supported by the repressive effect of CG methylation on gene expression.

@@

\begin{figure*}[t]
    \begin{center}
        \includegraphics[width=.8\textwidth]{figures/Fig_workflow.png}
    \end{center}
    \caption{\small \textbf{Overview of the functionalities implemented in the \texttt{scbs} package.}
        \texttt{scbs prepare} parses methylation files produced by common bisulfite sequencing mappers and stores their contents in a compressed format optimised for efficient access to genomic intervals.
        \texttt{scbs} also produces cell-wise summary statistics and quality plots (here: average methylation around the transcription start site) that are used to detect low-quality cells.
        These cells can be discarded with \texttt{scbs filter}.
        To obtain a methylation matrix, similar to the count matrices used in scRNA-seq, the user must first decide in which genomic intervals methylation should be quantified.
        The user may either provide genomic regions of \emph{a priori} interest, or they may choose to discover VMRs (variably methylated regions) in the data with \texttt{scbs scan}.
        The resulting methylation matrix can then be used for downstream analysis such as cell clustering and dimensionality reduction.
        Identified groups of cells (e.~g. clusters) can then be scanned for DMRs with \texttt{scbs diff}.
    }
    \label{fig:workflow}
\end{figure*}

We have implemented the approach just described in a Python package with a command line interface, called \texttt{scbs}, which also offers a number of other functionalities for the analysis of scBS data (Figure \ref{fig:workflow}).
The starting point of such an analysis are methylation files generated by tools such as Bismark \citep{bismark} or methylpy \citep{methylpy}.
For each cell, these tools produce a text file that lists the methylation status of all CpG sites.
Since it is inconvenient to work with hundreds or thousands of text files, \texttt{scbs} provides the \texttt{prepare} command which parses these methylation files and stores their content in a compressed format that enables efficient access to all CpG sites of the genome.
\texttt{scbs prepare} is flexible and accepts all tabular input formats including Bismark .cov-files, methylpy .allc-files, or custom user-defined formats.

@@

After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{scbs scan} command.
This produces a BED file that lists the genome coordinates of VMRs.
Similarly, the user can identify DMRs after quality control with \texttt{scbs diff} using a text file listing the cells’ group assignment as input.
The command produces a BED file listing the coordinates of the determined DMRs together with the t- statistic, the corresponding cell group, and the adjusted p-value.
To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells.
The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells.
We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells.
The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.

@@

Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably methylated regions (VMRs).
By re-analyzing two published data sets, we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data.
As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for smaller data sets with a limited number of cells.
In addition to this improvement of global signal-to-noise ratio (with regards to cell dissimilarity quantification), we also show that the determination of VMRs improve biological interpretability by pinpointing the regions that actually change and hence have a putative regulatory role.

@@

In addition, by pinpointing exact regions of interest, the VMR functionality provides useful input for genomics-style analyses.
Depending on the research question at hand, individual VMRs can be related to nearby genomic features such as gene bodies or known regulatory elements, tested for methylation differences between groups of cells, or subjected to gene ontology enrichment or motif enrichment.

@@

\begin{figure}
    \begin{center}
        \includegraphics[width=0.9\columnwidth]{figures/SFig_benchmark.png}
    \end{center}
    \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of neural subtypes of the murine frontal cortex}.
    Our methods were benchmarked on single-cell methylomes from \citet{luo2017single} as described in Fig.~\ref{fig:score}.
    }
    \label{fig:score_luo}
\end{figure}

@@

S.\ A.\ and L.\ P.\ M.\ K.\ acknowledge funding by the Klaus Tschira Foundation.
L.\ P.\ M.\ K., M.\ B., L.\ K., S.\ C.\ and A.\ M.-V.\ acknowledge funding from the European Commission via ERC grant 771376.
We thank Alexey Uvarovskii for code refactoring and helpful suggestions on the source code.

\phantomsection
\section{Methods}

@@

Whenever a region is not covered by any read in a cell, the corresponding data entry in the input data matrix for PCA will be missing.
The standard approach to calculate PCA (commonly done using the IRLBA algorithm, \citep{Baglama2005}) is not suited to deal with missing data.
We circumvent this issue by simply using the PCA itself to impute the missing value.

@@

Our iterative imputation strategy is now simply the following: we first replace all missing values in the row-centered input matrix $A$ with zeroes and perform the (truncated) PCA.
Then, we use the PCA predictions for the missing values, i.e., the $a_{ij}=\sum_r x_{ir} r_{jr}$, as refined stand-ins for the missing values in $A$ and run PCA once more.
This can now be iterated until convergence, which is typically achieved after at most 4--5 rounds of PCA.

\subsection{Re-analysis of published scBS datasets}
To re-analyze scBS data sets by \citet{luo2017single} and \citet{argelaguet2019gastru}, the genome was first annotated with genomic intervals that were either obtained by tiling the genome into non-overlapping 100~kb tiles, or by acquiring VMRs with \texttt{scbs scan} using the current default options (bandwidth $=2000$, stepsize $=10$, variance threshold $= 2\%$).
For all regions (VMRs and 100~kb tiles), the mean methylation fractions and the shrunken means of residuals were obtained with \texttt{scbs matrix}.
Regions containing less than four measured CpGs, or that were measured in less than 10\% of cells, were discarded.
To select $n$ intervals for further analysis, the top $n$ intervals were picked based on the number of cells that the region was measured in.
